---
title: " Group supervision meeting October 2025"
author: "Taps"
editor: visual
format: 
  html:
    colorlinks: true
    toc: true
    toc-depth: 3
    code-fold: true
---

-   **AGENDA**

    -   Discuss preliminary results - the discrepancies between LMM results and post-hoc analysis

**Title: Managing bush encroachment on a holistically managed rangeland in a semi-arid savanna**

**Research objectives:**   

i\. To determine the efficacy of selected bush encroachment management strategies including woody plant thinning, targeted herbicide application, browsers (goats), and prescribed fire, on seedling density, sapling recruitment, and resprouting dynamics.

ii\. To evaluate the effects of bush encroachment management strategies on above-ground grass biomass, grass species richness, and diversity at encroached sites.

**Research hypotheses**

a\) Combining woody plant thinning, prescribed fire, and goat browsing (TFB) will result in a substantial reduction in seedling and sapling densities. The impact of these combined management strategies is expected to be more pronounced in unfenced plots.

b\) Integrating woody plant thinning, targeted herbicide application, and prescribed fire (THF) on cut stumps would result in a substantial reduction in resprouts, and the decrease will be consistent across fenced and unfenced rangelands.

c\) The above-ground grass biomass will  increase across all treatments where tree thinning is done  (due to increased light penetration and water availability (Ludwig et al., 2004b). However, integrating woody plant thinning, prescribed fire, and targeted herbicide application (THF) will substantially increase grass above-ground biomass, with fenced plots experiencing a significant increase.

d\) Grass species richness is expected to increase in treatments where woody plants are thinned, as some grass species may have been excluded from the closed canopies. Although grass species richness and Shannon-Weiner index will significantly increase in the THF treatment, this increase will be more pronounced in across unfenced plots.

-    

    -   

        # **Data structure**

        str(strt_comparison2)

    ```         
    $ Site      : chr [1:188] "A" "A" "A" "A" ...   
    $ Plot      : num [1:188] 1 1 1 1 1 1 1 1 2 2 ...   
    $ Subplot   : chr [1:188] "Z1" "Z1" "Z2" "Z2" ...   
    $ Treatment : Factor w/ 5 levels "C", "F", "TF", "TFB", "THF"  
    $ Fencing   : Factor w/ 2 levels "Fenced", "Unfenced"   
    $ Year      : num [1:188] 2024 2025 2024 2025 2024 ...  
    $ Seedlings : int [1:188] 44 50 66 119 32 42 100 58 65 83 ...  
    $ density_ha: num [1:188] 733 833 1100 1983 533 ...   
    $ Period    : Factor w/ 2 levels "Pre-treatment","Post-Treatment"
    ```

    Response variable - Change in seedling density

    Fixed effects - Treatment, Fencing

    Random effects - Site, Plot

# [PRELIMINARY RESULTS]{.underline}

# a) Test efficacy of treatment and fencing on seedling density

```{r, warning=FALSE, message=FALSE}

library(tidyverse)
library(vegan)
library(multcompView)
library(patchwork)
library(lmerTest)
library(Matrix)
library(lme4)
library(emmeans)
library(here) ## MOST IMPORTANT NOT TO FORGET THIS ONE in quarto
library(robustbase)
library(sjPlot)
library(flextable)
library(officer)
library(glmmTMB)
library(DHARMa)   # model diagnostics GLMM
library(performance)  # model diagnostics
library(car)
library(openxlsx)
library(broom.mixed) #convert model objects to data frames
library(corrplot)
library(robustlmm) # for robust regression analysis
library(boot)  # For bootsrapping
library(DescTools)
library(effectsize) # For easy centering
library(marginaleffects)
library(effects)
library(ggeffects)

theme_beautiful <- function() {
  theme_bw() +
    theme(
      text = element_text(family = "Helvetica"),
      axis.text = element_text(size = 8, color = "black"),
      axis.title = element_text(size = 8, color = "black"),
      axis.line.x = element_line(size = 0.3, color = "black"),
      axis.line.y = element_line(size = 0.3, color = "black"),
      axis.ticks = element_line(size = 0.3, color = "black"),
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), units = , "cm"),
      plot.title = element_text(
        size = 8,
        vjust = 1,
        hjust = 0.5,
        color = "black"
      ),
      legend.text = element_text(size = 8, color = "black"),
      legend.title = element_text(size = 8, color = "black"),
      legend.position = c(0.9, 0.9),
      legend.key.size = unit(0.9, "line"),
      legend.background = element_rect(
        color = "black",
        fill = "transparent",
        size = 2,
        linetype = "blank"
      )
    )
}

## LOAD DATA

SapF <- read_csv(here("DATA/March2025/WoodyPC4.csv"))

## create seedling, sapling, trees and cut-stump row
SapF <- SapF %>% 
mutate(
woody_cat = case_when(
Woody_class == "Cut stump"          ~ "Cut stump",
between(`Max_height(m)`, 0.05, 0.50)~ "Seedlings",
between(`Max_height(m)`, 0.51, 1.49)~ "Saplings",
between(`Max_height(m)`, 1.5, 21.0) ~ "Trees",
TRUE                             ~ NA_character_
),
Treatment = factor(Treatment),
 Fencing = factor(Fencing)
)

 
#Filter SEEDLINGS  
Seedlings <- SapF %>% 
  filter(woody_cat == "Seedlings", Year %in% c(2024, 2025),
         !Treatment %in% c("TFB")) %>% 
  count(Site, Plot, Subplot, Treatment, Fencing,Year, name = "Seedlings") %>% 
  mutate(density_ha = Seedlings * 10000 / 600)     # convert to ha⁻¹

#comparing at treatment level

Seed_treat <- Seedlings %>% 
  group_by(Site, Plot, Subplot, Treatment, Fencing, Year) %>% 
  summarise(mean_dens_ha = mean(density_ha), .groups = "drop")  


#sort pre and post treatment
strt_comparison2 <- Seedlings %>%
  filter(Year %in% c(2024, 2025)) %>%
  mutate(Period = ifelse(Year == 2024, "Pre-treatment", "Post-treatment"))

#reorder so that pre-treatment appears first then post treatment second on the plots
strt_comparison2 <- strt_comparison2 %>%
  mutate(Period = factor(Period, levels = c("Pre-treatment", "Post-treatment")))

##Pivot the two years side‑by‑side and compute Δ seedling density
Seedlings_Delta <- Seed_treat %>% 
  pivot_wider(names_from  = Year,
              values_from = mean_dens_ha,
              names_glue  = "dens_{Year}") %>% 
  mutate(delta_Seeddens = dens_2025 - dens_2024) 

#GLMM to test effect of treatment * fencing on seedling density###################

# Make "Unfenced" the reference level 

class(Seedlings_Delta$Fencing)  # Likely "character" or "ordered factor"

# Convert to unordered factor explicitly
Seedlings_Delta$Fencing <- factor(Seedlings_Delta$Fencing, ordered = FALSE)

# Verify
levels(Seedlings_Delta$Fencing)  # Should show "Open" "Closed" (or vice versa)

# Set "Unfenced" as the reference level 
Seedlings_Delta$Fencing <- relevel(Seedlings_Delta$Fencing, ref = "Unfenced")


# Site and plot as (crossed) random effects
Seedl3 <- glmmTMB(delta_Seeddens ~ Treatment * Fencing + (1|Site),                data = Seedlings_Delta)


# check if model converged
#performance::check_convergence(Seedl3) # TRUE the model converged

# MODEL performance
#performance::check_model(Seedl3)

# Shapiro-Wilk Test for Normal distribution of residuals
 #shapiro.test(resid(Seedl3))  #p < 0.05 → residuals are non-normal.
# result p = 0.7431, hence residuals are normal. We have sufficient evidence to conclude that the sample data comes from a population that is normally distributed

## create a Q-Q plot of your residuals: 
 #qqnorm(resid(Seedl3)) 
 #qqline(resid(Seedl3))


##B. Levene’s Test for Homoscedasticity
# Bin fitted values into 3-5 groups
 #fitted_binned <- cut(fitted(Seedl3), breaks = 5)
 #leveneTest(resid(Seedl3) ~ fitted_binned)  # p < 0.05 → unequal variance (heteroscedasticity)
# results p = 0.918 equal variance 

# Assumption on Homoscedasticity

 # check homoscedasticity by creating a fitted value vs. residual plot. The scatterplot should show the fitted values of the model vs. the residuals of those fitted values.

# Line should be flat and horizontal if the homoscedasticity assumption is met.

#Dealing with heteroscedasticity (unequal variance) - log transforming the response variable, weighted regression (assigning weights to each data point based on the variance of its fitted value).

# results of the model
#summary(Seedl3)

```

Seedl4 \<- lmer(delta_Seeddens \~ Treatment \* Fencing + (1\|Site/Plot),\
data = Seedlings_Delta) \## did not converge

Seed5 \<- lmer(delta_Seeddens \~ Treatment \* Fencing + (1\|Site)+ (1\|Plot), data = Seedlings_Delta1) \## did not converge

Seedl3 \<- lmer(delta_Seeddens \~ Treatment \* Fencing + (1\|Site), data = Seedlings_Delta1)

-Before interpreting results from the model it is indispensible to :

\#**check if model converged**

performance::check_convergence(Seedl3) \# TRUE the model converged

\# **check model performance**

performance::check_model(Seedl3)

# Preliminary results from the model

```{r, warning=FALSE, message=FALSE}
#check if model converged

performance::check_convergence(Seedl3) # TRUE the model converged

# check model performance

performance::check_model(Seedl3)

# results of the model: change in seedling density  
summary(Seedl3)
```

**Post-hoc analysis for seedling density**

```{r, warning=FALSE, message=FALSE}
# Tukey HSD pairwise comparisons
treat_comparisons <- emmeans(Seedl3, specs = pairwise ~ Treatment | Fencing, adjust = "tukey")
summary(treat_comparisons$contrasts)


```

**Visualising results**

```{r, warning=FALSE, message=FALSE, fig.cap="Figure 1. Seedling density substantially reduced by TF in fenced plots at SHR five months post treatment."}
### Marginal effects for treatment * fencing on seedlings
Seeden <- ggpredict(Seedl3, terms = c("Treatment", "Fencing"))
plot(Seeden) +  #Store the plot in an object for saving 
  labs(y = "Change in seedling density per ha",
       x = "Treatment") +
  theme_beautiful()+ 
  ggtitle(NULL)+
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5)+
  theme(
    axis.title = element_text(size = 14),      # Axis titles
    axis.text = element_text(size = 12)        # Axis tick labels
  )
```

**Interpretation of Results**

1.  TF and fencing were associated with a significant reduction in seedling density (p = 0.0147). Treatment TF and fencing were highly effective at reducing the seedling proliferation.

2.  Post-hoc analysis reveal that THF and TF treatments in fenced plots significantly reduced seedling density when compared to the the baseline (p \< 0.05). 

# b) Test efficacy of treatment and fencing on sapling density

```{r, warning=FALSE, message=FALSE}
#Filter SAPLINGS  
Saplings <- SapF %>% 
  filter(woody_cat == "Saplings", Year %in% c(2024, 2025),
         !Treatment %in% c("TFB")) %>% 
  count(Site, Plot, Subplot, Treatment, Fencing,Year, name = "Saplings") %>% 
  mutate(density_ha = Saplings * 10000 / 600)     # convert to ha⁻¹

#comparing at treatment level

Sap_treat <- Saplings %>% 
  group_by(Site, Plot, Subplot, Treatment, Fencing, Year) %>% 
  summarise(mean_dens_ha = mean(density_ha), .groups = "drop")  

#Summary stats for saplings 
Saplsummary_stats <- Saplings %>%
  group_by(Treatment, Year) %>%
  summarise(
    N = n(),                                   # number of observations per treatment
    mean_density = mean(density_ha, na.rm = TRUE),
    sd_density = sd(density_ha, na.rm = TRUE)
  ) %>%
  ungroup()


#sort pre and post treatment
sptrt_comparison2 <- Saplings %>%
  filter(Year %in% c(2024, 2025)) %>%
  mutate(Period = ifelse(Year == 2024, "Pre-treatment", "Post-treatment"))

##reorder so that pre-treatment appears first then post treatment second on the plots
sptrt_comparison2 <- sptrt_comparison2 %>%
  mutate(Period = factor(Period, levels = c("Pre-treatment", "Post-treatment")))

#Pivot the two years side‑by‑side and compute Δ sapling density
Saplings_Delta <- Sap_treat %>% 
  pivot_wider(names_from  = Year,
              values_from = mean_dens_ha,
              names_glue  = "dens_{Year}") %>% 
  mutate(delta_Sapsdens = dens_2025 - dens_2024) 


# Make "Unfenced" the reference level 

class(Saplings_Delta$Fencing)  # Likely "character" or "ordered factor"

# Convert to unordered factor explicitly
Saplings_Delta$Fencing <- factor(Saplings_Delta$Fencing, ordered = FALSE)

# Verify
levels(Saplings_Delta$Fencing)  # Should show "Open" "Closed" (or vice versa)

# Set "Fenced" as the reference level 
Saplings_Delta$Fencing <- relevel(Saplings_Delta$Fencing, ref = "Unfenced")

#### Convert character variables to factors
Saplings_Delta$Treatment <- as.factor(Saplings_Delta$Treatment)
Saplings_Delta$Fencing <- as.factor(Saplings_Delta$Fencing)


# LMM for sapling
Sapl5 <- lmer(delta_Sapsdens ~ Treatment * Fencing + (1|Site),  
                 data = Saplings_Delta)
summary(Sapl5)
```

**Post hoc analysis for sapling density**

```{r,warning=FALSE, message=FALSE}
# Tukey HSD pairwise comparisons
Sptreat_comparisons <- emmeans(Sapl5, specs = pairwise ~ Treatment | Fencing, adjust = "tukey")
summary(Sptreat_comparisons$contrasts)

```

**Visualising results**

```{r,warning=FALSE, message=FALSE, fig.cap="Figure 2. Sapling density substantially reduced by THF in unfenced plots at SHR five months post treatment."}

#marginal effects
Sapden <- ggpredict(Sapl5, terms = c("Treatment", "Fencing"))
plot(Sapden) +     
  labs(y = "Change in sapling density per ha",
       x = "Treatment") +
  theme_beautiful()+ 
  ggtitle(NULL)+
  theme(
    axis.title = element_text(size = 14),      # Axis titles
    axis.text = element_text(size = 12)        # Axis tick labels
  )+
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5)

```

**Results interpretation**

-   LMM analysis reveal that THF substantially reduced sapling density in unfenced plots (p \< 0.05) similarly the post-hoc analysis show that THF in unfenced plots significantly reduced sapling density (p \< 0.05). Furthermore, the post-hoc analysis reveal that TF and THF in fenced plots strongly reduced sapling density (p \< 0.05).

# c) Test efficacy of treatment and fencing on grass Shannon-Weiner index

```{r,warning=FALSE, message=FALSE, fig.cap="Figure 3. Grass Shannon-Weiner diversity substantially reduced by THF in unfenced plots at SHR five months post treatment."}

#load data
Grasses <- read_csv(here("DATA/March2025/GrassesCombinedCleaned3.csv"))

# Calculate Shannon-Wiener Diversity Index at Site and Plot level
GrSWeiner <- Grasses%>%
  filter(!is.na(Species_name),
         Year %in% c(2024, 2025),!Treatment %in% c("TFB"))%>%   # keep only pre/post years
  group_by(Site, Plot, Subplot, Treatment, Year, Fencing, Species_name)%>%
  summarise(Spp_count = n(), .groups = "drop" )

## Calculate Shannon-Wiener Diversity Index at Site and Plot level
grSWdiversity <- GrSWeiner %>%
  group_by(Site, Plot, Subplot,Treatment, Year, Fencing) %>%                   # Group by Site and Plot
  summarise(
    Shannon_Diversity = -sum((Spp_count / sum(Spp_count)) * (Spp_count / sum(Spp_count))),
    .groups = "drop")%>%
  mutate(Period = ifelse(Year == 2024, "Pre-treatment", "Post-treatment"))

##reorder so that pre-treatment appears first then post treatment second on the plots
grSWdiversity <- grSWdiversity %>%
  mutate(Period = factor(Period, levels = c("Pre-treatment", "Post-treatment")))

###### DELTA SHANNON-WEINER DIVERSITY INDEX
Gdiv <- Grasses%>%
  filter(!is.na(Species_name),
         Year %in% c(2024, 2025), !Treatment %in% c("TFB"))%>%   # keep only pre/post years
  group_by(Site, Plot, Subplot, Treatment, Year, Fencing, Species_name)%>%
  summarise(Spp_count = n(), .groups = "drop" )

# Calculate Shannon-Wiener Diversity Index log transforming
Sdiversity_data <- Gdiv %>%
  group_by(Site, Plot, Subplot,Treatment, Year, Fencing) %>%   # Group by Site and Plot
  summarise(
    Shannon_Diversity = -sum((Spp_count / sum(Spp_count)) * (Spp_count / sum(Spp_count))),
    .groups = "drop"
  )

#  Pivot the two years side‑by‑side and compute Δ‑ grass Shannon-Weiner diversity
SW_Delta <- Sdiversity_data %>% 
  pivot_wider(names_from  = Year,
              values_from = Shannon_Diversity,
              names_glue  = "sw_{Year}") %>% 
  mutate(delta_SW = sw_2025 - sw_2024)   

### Convert character variables to factors
SW_Delta$Treatment <- as.factor(SW_Delta$Treatment)
SW_Delta$Fencing <- as.factor(SW_Delta$Fencing)

# Make "Unfenced" the reference level 

class(SW_Delta$Fencing)  # Likely "character" or "ordered factor"

# Convert to unordered factor explicitly
SW_Delta$Fencing <- factor(SW_Delta$Fencing, ordered = FALSE)

# Verify
levels(SW_Delta$Fencing)  # Should show "Open" "Closed" (or vice versa)

# Set "Fenced" as the reference level 
SW_Delta$Fencing <- relevel(SW_Delta$Fencing, ref = "Unfenced")

### Convert character variables to factors
SW_Delta$Treatment <- as.factor(SW_Delta$Treatment)
SW_Delta$Fencing <- as.factor(SW_Delta$Fencing)


#LMM
GrasD <- lmer(delta_SW ~ Treatment * Fencing + (1|Site),  
                   data = SW_Delta)

#check if model converged

performance::check_convergence(GrasD) # TRUE the model converged

# check model performance

performance::check_model(GrasD)

# results of the model: change in seedling density  

summary(GrasD)
```

**Post-hoc analysis**

```{r,warning=FALSE, message=FALSE}
# Tukey HSD pairwise comparisons
Grass_diversity <- emmeans(GrasD, specs = pairwise ~ Treatment | Fencing, adjust = "tukey")
summary(Grass_diversity$contrasts)


```

**Visualisation grass diversity plot**

```{r,warning=FALSE, message=FALSE, fig.cap="Figure 3. Grass Shannon-Weiner diversity index was not affected by treatment and fencing at SHR five months post treatment."}
#marginal effects
preds <- ggpredict(GrasD, terms = c("Treatment", "Fencing"))
 plot(preds) +  
  labs(y = "Change in Shannon-Weiner Diversity",
       x = "Treatment") +
  theme_beautiful()+ 
  ggtitle(NULL)+
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5)+
  theme(
    axis.title = element_text(size = 14),      # Axis titles
    axis.text = element_text(size = 12)        # Axis tick labels
  )

```

**Results interpretation**

-   Treatment and fencing (F, TF, THF) did not significantly impact grass Shannon-Weiner diversity index (p \> 0.05) . the Control in unfenced plots had a strong positive effect (p \<0.05).

-   Post-hoc analysis show no significant difference for all the treatments when compared to the control (p \> 0.05).
